\documentstyle[12pt,letterpaper]{letter}
\begin{document}
\pagestyle{empty}

\signature{Matt Challacombe}

\address{
Dr.~Matt Challacombe \\
(mchalla@lanl.gov) \\
Theoretical Division \\
Group T-12, MS B268 \\
Los Alamos National Laboratory \\
Los Alamos, NM 87545 }

\begin{letter}{
Professor Donald Levy, Editor\\
Journal of Chemical Physics\\
Department of Chemistry\\
5735 S. Ellis Ave\\
The University of Chicago\\
Chicago, IL 60637}

\opening{Dear Professor Levy,}

Enclosed, please find two copies of 
{\em Linear Scaling Computation of the Fock Matrix.~VI.~Data Parallel Computation 
of the Exchange-Correlation Matrix} by Gan and Challacombe,
which I am submitting for publication in the Journal of Chemical Physics (JCP). This 
article builds 
on our last JCP publication {\em Linear Scaling Computation of the Fock Matrix.~V.~Hierarchical Cubature
for numerical integration of the Exchange-Correlation Matrix}.

This article is more than just a numerical recipe, establishing a strong connection between linear 
scaling methods and methods that are efficient in parallel. In particular, this contribution 
makes the following points:

\pagebreak


\begin{itemize}
\item The properties of quantum locality that allow linear scaling should also lead to 
      data locality and scalable parallel algorithms.  This is an important and completely 
      unexplored aspect of the linear scaling paradigm.

\item The correspondence between quantum locality and data locality requires a new approach
      that we have taken with our Hierarchical Cubature method. 
      The results achieved here for fine grained massive parallelism speak for themselves.

\item A novel and general load balancing scheme  well suited to many aspects of quantum chemistry is developed.
\end{itemize}

The Journal of Chemical Physics has a rich history of publishing work on numerical methods for 
Electronic Structure Theory.  More and more, the success of these numerical methods are
determined by their ability to map efficiently onto parallel platforms.  This
is clearly reflected in a number of recent JCP articles on parallel methods, notably 
{\em Parallel filter diagonalization: A novel method to resolve quantum states in dense spectral regions} and
{\em Multigrid methods for classical molecular dynamics simulations of biomolecules}.  

With this in mind, I hope you will consider this submission for publication in the Journal of Chemical Physics.
Suggestions for reviewers with expertise in both parallel methods and Quantum Chemistry include:
Michael Colvin at LLNL (colvin2@llnl.gov), Emily Carter at UCLA (eac@chem.ucla.edu),
Peter Taylor (ptaylor@ucsd.edu) or Kim Baldridge (kimb@sdsc.edu) at San Diego,
Bill Goddard at CalTech (wag@wag.caltech.edu), Tom Furlani at SUNY (furlani@chem.buffalo.edu), 
and Mark Gordon at Ames (mark@si.fi.ameslab.gov).  


\closing{With best regards,}
\end{letter}
\end{document}

