\documentstyle[12pt,letterpaper]{letter}
\begin{document}
\pagestyle{empty}

\signature{Matt Challacombe}

\address{
Dr.~Matt Challacombe \\
(mchalla@lanl.gov) \\
Theoretical Division \\
Group T-12, MS B268 \\
Los Alamos National Laboratory \\
Los Alamos, NM 87545 }

\begin{letter}{
Professor Ernest R. Davidson, Associate Editor\\
Department of Chemistry\\
University of Washington,\\
Box 351700/Bagley 303A,\\
Seattle, WA 98195-1700, USA.
}

\opening{Dear Professor Davidson,}

Enclosed, please find two revised copies (A2.11.022) of 
{\em Linear Scaling Computation of the Fock Matrix.~VI.~Data 
Parallel Computation 
of the Exchange-Correlation Matrix} by Gan and Challacombe, which has been
reviewed by The Journal of Chemical Physics.

The changes that have been made are as follows:
\pagebreak

\begin{itemize}
\item A general Equal Time partitioning scheme has been 
outlined in the Appendix. This scheme has been included in response to
the first comment of the reviewer. In this scheme, we discuss how the original
power-of-two condition for the number of processors 
can relaxed so that Equal Time partition may be obtained with
an arbitrary number of processors.

\item In response to the second comment of the reviewer,
we have performed $\alpha$-pinene calculations with the 6-311G(df,p) basis
set to show the applicability of parallel HiCu to a small system but with
a large basis set. The results have been included in the manuscript. 
The abstract and conclusion have been slightly changed to reflect 
the new results. 

\item We acknowledged the valuable comments of the reviewer.

\end{itemize}

We thank the reviewer for valuable comments.

\closing{With best regards,}
\end{letter}
\end{document}

